Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-17 article.tex
format code lines to prevent them running off right edge of article -- add linebreaks in convenient places

instance (Functor m, Switch m) => Switch (StateT s m) where
  switch (StateT f) = StateT (\s -> fmap (const ((), s)) . switch $ f s)

@@

*Main> map (runParser opencurly) [";abc\n  {def", "{def", "(def", "}def"]
[Just ('{',"def"),Just ('{',"def"),Nothing,Nothing]
\end{verbatim}

@@

*Main> runParser (throwError "no backtracking!" <|> satisfy (== 'b')) "babc"
Left "no backtracking!"
\end{verbatim}
We'll take advantage of this backtracking restriction to produce accurate error
messages.

@@

instance MonadError e m => MonadError e (StateT s m) where
  throwError      =  lift . throwError
  catchError m f  =  StateT (\s -> catchError (runStateT m s) 
                                              (\e -> runStateT (f e) s))

@@

And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
*Main> runParser woof (countLineCol "{define fgh {lambda {f x} (f x)}}")
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") [ASymbol "x"]])],
       []))
\end{verbatim}
Here's the proof showing that the parser correctly reports all of the errors 
covered by our spec:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

*Main> p "()"
Left ("application: missing operator",('(',1,1))

>>>>

instance (Functor m, Switch m) => Switch (StateT s m) where
  switch (StateT f) = StateT g
    where 
      g s = fmap (const ((), s)) . switch $ f s

@@

*Main> runParser opencurly ";abc\n{def"
Just ('{', "def")

*Main> runParser opencurly "{def"
Just ('{', "def")

*Main> runParser opencurly "(def"
Nothing

*Main> runParser opencurly "}def"
Nothing
\end{verbatim}

@@

*Main> runParser (throwError "no!" <|> satisfy (== 'b')) "babc"
Left "no!"
\end{verbatim}
We'll take advantage of this backtracking restriction to produce accurate error
messages.

@@

instance MonadError e m => MonadError e (StateT s m) where
  throwError      =  lift . throwError
  catchError m f  =  StateT g
    where
      g s = catchError (runStateT m s) 
                       (\e -> runStateT (f e) s)

@@

And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 chars
  where chars = not1 $ character '\n'

@@

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (f . chr)
        f = flip elem (['a' .. 'z'] ++ ['A' .. 'Z'])
\end{verbatim}

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

*Main> p "{define fgh {lambda {f x} (f x)}}"
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") 
                                           [ASymbol "x"]])],
       []))
\end{verbatim}
Here's the proof showing that the parser correctly reports all of the errors 
covered by our spec:
\begin{verbatim}
*Main> p "()"
Left ("application: missing operator",('(',1,1))
